Boundary Labeling for annotated documents

### 2018-04-03 LaTeX/Paper.tex
Paper: Re-Wrote Chapter 3 to add separate sections for GUI, Drawing Area and the Routing Algorithm. Program: No changes.

\documentclass[11pt,a4paper]{vutinfth}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage[boxed,linesnumbered]{algorithm2e}
\usepackage{color}
\usepackage{subcaption}
\usepackage{amsthm}

%TODO: CHANGE!
\setauthor{}{Jakob Klinger}{}{male}
\setadvisor{Ass.Prof. Dipl.-Inform. Dr.rer.nat}{Martin Nöllenburg}{}{male}%{Pretitle}{Forename Surname}{Posttitle}{male}
\setfirstassistant{Univ.Ass.}{Fabian Klute}{ M.Sc., B.Sc.}{male}
%Ass.Prof. Dipl.-Inform. Dr.rer.nat. Martin Nöllenburg
% University Assistant, M.Sc., B.Sc. Fabian Klute

@@

\setthesis{bachelor}
\setcurriculum{Bachelor's programme Software \& Information Engineering}{Bachelorstudium Software \& Information Engineering} %TODO!

@@

\section{Terminology and Fundamentals} 
Boundary Labeling (or equivalent concepts) can be applied to a space with different geometry or more dimensions, but this thesis will only concern itself with two-dimensional, Euclidean space.
To easily reference important concepts, some additional terminology will be introduced as well. (See Fig.~\ref{fig:term} for a visual explanation)

@@

\emph{Labels} hold additional information and are represented as boxes containing this information. They always have a \emph{port}, a special point on the label's border which will be defined later. Labels are usually placed in the \emph{label area} which is a rectangular area designated to hold labels. It is located next to of the bigger, rectangular \emph{text area}, which contains the document's text and all \emph{sites}, the points or objects that a label's information refers to. If multiple label areas exist on different sides of the text area, we speak of \emph{multi-sided labeling}, otherwise we speak of \emph{one-sided labeling}. We will be using one-sided labeling in our implementation. Some space was left in between the text and label area, to make connecting sites to their labels easier, which we will call the \emph{routing area}. The site and the label are connected via a \emph{leader}, a polyline that can be further classified by looking at the orientation of its segments: \emph{O-Segments} run orthogonally to the border of the label area. \emph{P-Segments} run parallel to the border of the label area, and as such must be combined with other segments for the leader to reach its destination. \emph{S-Segments} are not required to have any particular orientation, and simply connect their start and ending points in a straight line.
The leader's name is created by combining the name of the segments - for example, the blue leader from Fig~\ref{fig:term} would be classified as an OPOPO-Leader.
The location where a leader connects to the label is called the \emph{port}. It can be restricted to pre-defined positions. %TODO: Re-insert monotonicity

@@

\chapter{The Algorithm}
%1 Zeile Einleitung hier einfügen - zb: Kapitelübersicht
\section{Problem Specification}
\label{sec:ProbSpec}
We limited the leaders to use exclusively O- and P-Segments, and banned them from passing through words. We also place labels as far up as possible to maximize the space remaining for  remaining placements. Additionally, a leader isn't allowed to be any longer as is necessary to connect a given Site with its label's port.

These restrictions are implemented as follows: We divided the text $T=\{W,H\}$ into separate lines $L=\{l_1 ,l_2 , ... ,l_n\}$ which are in turn made up of words $w \in W$ and whitespace $h \in H$, which alternate, starting and ending with a word, which creates a sequence $l=(w_1,h_1,w_2, ... , h_{m-1},w_m)$, where $m$ is the number of words in $l$. The width of each line is equal to the text area's total width in characters $(width)$, which means that $m \leq \frac{width}{2}$. The remaining space between lines is called $S=\{s_1, ... s_{n+1}\}$, with $n$ being the number of lines in $L$. For each $l_i \in L$,  $s_i$ is the space directly above that line. All $s \in S$ are the same height and width, even $s_1$ and $s_{n+1}$, which only have lines on one of their sides. The $s \in S$ and $l \in L$ are arranged in alternating order in the text $T=\{s_1, l_1, s_2, l_2, ..., s_n, l_n, s_{n+1}\}$.  For each word $w \in W$, we define $R(w)$  as its bounding rectangle, which marks the space leaders aren't allowed to cross. For a graphical representation see Fig.~\ref{fig:wbound}.

@@

Since we only allowed the Usage of O- and P-Segments in leaders, the only way for a leader to cross through a line of text is with a P-Segment through whitespace, whereas O-Segments are only usable between lines. Therefore, we can create a routing graph $G=\langle V,E\rangle$ whose Edges $E$ reflect the legal paths a leader can take within $T$.
For each line $l_i \in L$ and each whitespace character $h \in l_i$, one vertex is placed in $s_i$ and $s_{i+1}$ each, located above or below the whitespace character, and with a maximum distance from each of its neighbouring lines. The start and end of each line are also assigned a pair of vertices each. Sites will be represented by the insertion of additional vertices in $S_i$ on a similar height as those next to whitespace of the same line and located directly above the center point of its words $w_j$ bounding rectangle $R(w_j)$. The edges between the vertices are all perfectly horizontal or vertical, and do neither intersect with any bounding rectangle, nor any vertices other than their starting and ending vertex. The resulting graph looks similar to Fig.~\ref{fig:rgraph}.
In the following, we define vertex $v_a \in s_i$ to be \emph{above} another vertex $v_b \in s_j$, if $i<j$. If $v_a$ and $v_b$ are adjacent to each other, $v_a$ is \emph{directly above} $v_b$. We will also define a vertex $v_b$ to be \emph{between} two vertices $v_a$ and $v_c$, if $v_a, v_b, v_c \in s$ $(s \in S)$ and it is impossible to create a path $P \subseteq s$ using only vertices from $s$ that leads from $v_a$ to $v_c$ without including $v_b$.
Furthermore,we will call the vertices closest to the text area's border to the routing area \emph{Border Vertices} $B \subset V$. They serve as a goal for the depth-first search, and are located at the end of each line in our implementation.

@@

We use a depth-first-search algorithm with the Graph $G$ and the set of sites $V_{ann} \subset V$ as input. The Algorithm prioritizes routing to not yet visited vertices located above the current vertex and terminates either when reaching the right text border or if all possible paths failed to reach the text border, and the resulting SP-path will be monotonically increasing. If the routing for any site $v_{ann} \in V_{ann}$ failed, the algorithm returns $\bot$ for that site. \change{To ensure that routings remain crossing-free, the most recently successfully routed leader $P_last$ is used as a ceiling to the available search space. This means that any vertex above the polyline created by the vertices in $P_last$ and the connection of $P_last$'s source with the text area's border opposite of the label area is forbidden to use. (See als Fig.~\ref{fig:ceil})} This path is split into two parts: The path within the graph $(P \subset V)$, leading from the source to the text area's border, and the OPO-Leader that connects the text area's border with the Label's port. The former is implemented as a list of vertices $V_{Path} \subset V$ the leader travels through, whereas the latter is determined by the rightmost vertex of $V_{Path}$, the Port, and the location of the P-Segment. (For an illustration in Pseudocode see Alg.~\ref{alg:DFS}.)

@@

%\begin{itemize}
% \item Computation time of graph-creation
% \item Computation time of the routing algorithm
%\end{itemize}
%~\\
%Furthermore, we will prove that:
%\begin{itemize}
%\item The algorithm generates only crossing-free routings
%\item The algorithm always returns a highest (define!) legal path
%
%NOTE: The algorithm keeps the annnotations in reading order
%\end{itemize}
%~\\
\begin{lemma}
	The creation of the routing graph is in O(n).
	\label{lem:GraphCreat}
\end{lemma}

\begin{proof}
	As discussed in Section~\ref{sec:ProbSpec}, each line holds at most $\frac{width}{2}$ words, with an appropriate amount of whitespace in between. Since $width$ is a constant value, the amount of vertices placed is only dependent on the number of lines in the text, therefore $|V|=O(n)$.
	Similarly, each vertex has at most 4 incident edges, representing the two possible placements of P- or O-Segments starting from that vertex, respectively. Therefore, $|E|=O(n)$. As both the creation of new vertices and edges can be done in a constant amount of time, the overall computation time is in $O(n)$.
\end{proof}

%As discussed in section~\ref{sec:ProbSpec}, we will be placing a pair of vertices per whitespace character, as well as one additional pair at both the start and end of each line, which amounts to $2\times|H|+4\times|L|$ vertices total. Together with the sites $I$, which are also represented as vertices, this results in the total number of vertices $|V|=|I|+2\times|H|+4\times|L|$. Additionally, each vertex in a pair is connected to its counterpart, which creates $(|V|-|I|)/2$ edges, representing all possible P-Segments. Finally, each edge placed in a space $s \in S$ above or below a line will be connected to each other, which is solved by connecting each vertex to the next unconnected vertex remaining in $s$. This leads to the creation of $|s|-1$ edges representing possible P-segments, which leads to a total of $|V|-|S|$ additional edges, making the total number of edges $|E|=(3\times|V|-2\times|S|-|I|)/2$.
%If we assume that the creation of edges and vertices both take a similar amount of time, we are looking at roughly $5/2 \times |V|$ operations, which means that the algorithm scales linearly with the amount of vertices placed, which in turn is directly proportional to the number of words $|W|$ in the text.
For a representation of the graph-generation algorithm in pseudocode, see Alg.~\ref{alg:GraphCreation}.

@@

%Next, we shall inspect the routing algorithm - since we use depth-first search, which is a well-known algorithm, the worst-case computation time is known to be $|V|+|E|$. As both $|V|$ and $|E|$ are in $O(n)$, so is our routing algorithm.

%To reach this time, we'd have to try and use every single edge in the graph, reaching each of the graph's vertices in the process. However, as our algorithm can't use any edges leading away from its target, our search space is reduced to only the vertices either in lower-numbered $s \in S$ (those ``above'' the current position) or those in the same $s$, but located between the source and the border vertices. If we also take into account that leaders aren't allowed to cross one another, this further limits our search space, removing all vertices above the previous leader's vertices.

\begin{lemma}
	The algorithm only creates crossing-free routings.
	\label{lem:CrosFree}
\end{lemma}
\begin{proof}
	As stated in Section~\ref{sec:AlgDesc}, we are forbidden to use any vertices above \change{the polyline created by the last successfully routed leader $(P_{last})$ and the direct connection of $P_{last}$'s source and the left text area's border. As this makes it impossible to construct a route that incorporates any vertex above or to the left of $P_{last}$, those two leaders cannot possibly cross each other. This is true for any two consecutive leaders, therefore each leader also cannot cross any of the leaders routed before $P_{last}$, as those are located on the other side of $P_{last}$, which contains only forbidden vertices.
	Therefore, our algorithm produces crossing-free routings.}
\end{proof}

\change{For the next part, we first need to define what we consider a \emph{highest legal path}: A \emph{legal path} is one that satisfies both the constraints of monotonicity and being crossing-free, whereas a \emph{highest path} reaches the highest possible border vertex reachable from a given source.}

\begin{lemma}
	Our algorithm only returns a highest legal path.
	\label{lem:High}
\end{lemma}
\begin{proof}
	The legality of any generated path is given, due to the algorithm only returning monotonically increasing paths, and Lemma~\ref{lem:CrosFree}. To show that there are no reachable border vertices above the path's highest reachable border vertex $b_{high}$, we will assume there exists a vertex $b_{top} \in B$ that is located above $b_{high}$ and legally reachable via the path $P_{top}$. As the algorithm prioritizes routing to vertices directly above, the only way for the path to $b_{top}$ to reach higher vertices than $b_{high}$ would be that either $P_{top}$ would generate crossings with previously routed leaders or violate the the monotonicity constraint, which both render $P_{top}$ illegal. Since we also cannot include more than one border vertex in our path, we conclude that such a path cannot exist.
\end{proof}

\begin{theorem}
	The routing problem can be solved with a worst-case time complexity of $O(n^2)$ for all sources.
\end{theorem}
\begin{proof}
	As part of the proof for Lemma~\ref{lem:GraphCreat}, we have shown that the total number of vertices $|V|$ is in $O(n)$. Since $S \subset V$, $|S|$ must too be in $O(n)$. 
	We have proven our algorithm's solutions to be correct in the Lemmas \ref{lem:CrosFree} and \ref{lem:High}.
	The performance of our routing algorithm is easily calculated, as depth-first-search has a worst-case-performance of $|V|+|E|$, which is in $O(n)$ for our problem, as $|E|$ is in $O(n)$ as well. (See Lemma~\ref{lem:GraphCreat})
	Therefore, our overall computation time comes out to $|S| \times O(n)$, or $O(n^2)$.
\end{proof}

\chapter{Implementation}
The program was written in Java 1.8.0u40, using JGraphT 1.0.1\cite{JGraphT} as graph library. No other external libraries were used.
The program itself can be found at~\cite{Program}.

\section{Architecture}
The program is divided into three parts. The GUI and program window only propagate user interaction to other parts of the program. It will be discussed further in the next section.
The main drawing area handles the visualization of the results, calculates where to place the text on-screen, and also generates the routing graph, as this can be done alongside the text placement.
The routing is done in a separate class, and can easily be switched out for another algorithm because of this.
%The routing itself is handled separately from this process in its own class, as the finished routing graph suffices here. This facilitates the implementation of additional routing algorithms and allows for easy switching. The only other notable part of the program is the GUI, which will be discussed in the next section.%Since we only want to create leaders that don't intersect with the text, the graph was created alongside the placement of the words on the canvas.%, as it was easy to extract measurements at this point. Due to some problems with Java's various methods of calculating linebreaks, this process was done completely by hand.
\section{GUI}
\label{sec:GUI}

The GUI of the program is kept simple, as the main goal was to demonstrate the routing algorithm. All customization options are located on the control panel above the drawing area, which contains the text, leaders and annotations. (See also Fig.~\ref{fig:ProgWindow})

The control panel's first three options determine the font the program uses, whether it is bold or italicized and its size. The fourth and last option allows the selection of different routing algorithms. After changing any value, the program will automatically be notified of the changes and react accordingly.

The drawing area itself can't be interacted with, it only serves to display the algorithm's results.
\begin{figure}
	\captionsetup{justification=centering, margin=0.75cm}
	\centering
	\includegraphics[scale=0.75]{ProgramWindowAnnotated.png}
	\caption{Picture of the routing program}
	\label{fig:ProgWindow}
\end{figure}

\section{Challenges}

Displaying the text on-screen without losing the position of the whitespace characters turned out to be harder than expected. While Java offers a class that divides any string of characters in separate lines, the resulting line object doesn't allow for it's text to be extracted, which makes it unusable for our purposes. This problem was solved by placing each word manually on the screen. For simplicity's sake, we only allowed line breaks between words, whenever a whitespace character was encountered.

Another issue was the visualization of the routing graph and the leaders. While JGraphT offers ways to display graphs, it wasn't possible to influence the placement of individual vertices. Our solution was forgoing these options and drawing the graph manually, assigning coordinates to each vertex on creation. Vertices are represented by small rectangles, whereas edges are represented with straight lines. To avoid confusion between crossing leaders and leaders that incorporate the same vertex in their path, but curve away from each other, all leaders are drawn with rounded corners.

@@

When the program starts, it first goes through the text word by word in reading order, placing each word $w \in W$  on-screen, and saving the content of each encountered annotation for later. As the words are placed, vertices are added to the graph after each word, and at the start and end of each line. Source vertices are inserted whenever an annotation is encountered, and contain a reference to their annotation in addition to the information contained in regular vertices.

@@

Once the graph has been generated, a list of all source vertices in reading order is passed to the selected routing algorithm, which then determines leader routing and annotation placement for each of them and returns the results to the main program, which then draws the results. (For details on the routing algorithm, see section~\ref{sec:AlgDesc}.)

If any of the parameters on the control panel (see section~\ref{sec:GUI}) are changed, the process re-starts from the beginning.
%As the Graph's vertices represent fixed locations on the canvas, they each have coordinates associated with them, with the vertices representing the sites containing extra information, such as references to the corresponding annotation and to the leader connecting the two (if existent).
%The additional information stored in the vertices is used for both routing and drawing the leaders: The routing was done via a depth-first search algorithm that prioritized vertices located further up if possible, and vertices to the right otherwise. After the algorithm terminates, it returns  a vertex-based graph walk describing the route from the site to the border of the text area alongside information about where to draw the OPO-segment in the buffer zone if successful, or a walk containing only the starting vertex, if no valid path was found. Using this information alongside the positional data stored in the vertices, the polyline representing the leader can be drawn.

%TODO: Schönheitsfeatures, etc.
%TODO: Code genauer durchlesen und erweitern!

% For demonstration purposes, we also included another algorithm that uses S-Leaders for this section.

>>>>

\documentclass[11pt,a4paper]{vutinfth}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{caption}
\usepackage{mathtools}
\usepackage{amssymb}
\usepackage[boxed,linesnumbered]{algorithm2e}
\usepackage{color}
\usepackage{subcaption}
\usepackage{amsthm}
\usepackage[hyperfootnotes=false]{hyperref}

\setauthor{}{Jakob Klinger}{}{male}
\setadvisor{Ass.Prof. Dipl.-Inform. Dr.rer.nat}{Martin Nöllenburg}{}{male}%{Pretitle}{Forename Surname}{Posttitle}{male}
\setfirstassistant{Univ.Ass.}{Fabian Klute}{ M.Sc., B.Sc.}{male}
%Ass.Prof. Dipl.-Inform. Dr.rer.nat. Martin Nöllenburg
% University Assistant, M.Sc., B.Sc. Fabian Klute

@@

\setthesis{bachelor}
\setcurriculum{Bachelor's programme Software \& Information Engineering}{Bachelorstudium Software \& Information Engineering}

@@

\section{Terminology and Fundamentals} 
Boundary Labeling (or equivalent concepts) can be applied to a space with different geometry or more dimensions, but this thesis will only concern itself with two-dimensional, Euclidean space.
To easily reference important concepts, some additional terminology will be introduced as well. (See Fig.~\ref*{fig:term} for a visual explanation)

@@

\emph{Labels} hold additional information and are represented as boxes containing this information. They always have a \emph{port}, a special point on the label's border which will be defined later. Labels are usually placed in the \emph{label area} which is a rectangular area designated to hold labels. It is located next to of the bigger, rectangular \emph{text area}, which contains the document's text and all \emph{sites}, the points or objects that a label's information refers to. If multiple label areas exist on different sides of the text area, we speak of \emph{multi-sided labeling}, otherwise we speak of \emph{one-sided labeling}. We will be using one-sided labeling in our implementation. Some space was left in between the text and label area, to make connecting sites to their labels easier, which we will call the \emph{routing area}. The site and the label are connected via a \emph{leader}, a polyline that can be further classified by looking at the orientation of its segments: \emph{O-Segments} run orthogonally to the border of the label area. \emph{P-Segments} run parallel to the border of the label area, and as such must be combined with other segments for the leader to reach its destination. \emph{S-Segments} are not required to have any particular orientation, and simply connect their start and ending points in a straight line.
The leader's name is created by combining the name of the segments - for example, the blue leader from Fig~\ref*{fig:term} would be classified as an OPOPO-Leader.
The location where a leader connects to the label is called the \emph{port}. It can be restricted to pre-defined positions.

@@

\chapter{The Algorithm}
%TODO:1 Zeile Einleitung hier einfügen - zb: Kapitelübersicht
\section{Problem Specification}
\label{sec:ProbSpec}
We limited the leaders to use exclusively O- and P-Segments, and banned them from passing through words. We also place labels as far up as possible to maximize the space remaining for  remaining placements. Additionally, a leader isn't allowed to be any longer as is necessary to connect a given Site with its label's port.

These restrictions are implemented as follows: We divided the text $T=\{W,H\}$ into separate lines $L=\{l_1 ,l_2 , ... ,l_n\}$ which are in turn made up of words $w \in W$ and whitespace $h \in H$, which alternate, starting and ending with a word, which creates a sequence $l=(w_1,h_1,w_2, ... , h_{m-1},w_m)$, where $m$ is the number of words in $l$. The width of each line is equal to the text area's total width in characters $(width)$, which means that $m \leq \frac{width}{2}$. The remaining space between lines is called $S=\{s_1, ... s_{n+1}\}$, with $n$ being the number of lines in $L$. For each $l_i \in L$,  $s_i$ is the space directly above that line. All $s \in S$ are the same height and width, even $s_1$ and $s_{n+1}$, which only have lines on one of their sides. The $s \in S$ and $l \in L$ are arranged in alternating order in the text $T=\{s_1, l_1, s_2, l_2, ..., s_n, l_n, s_{n+1}\}$.  For each word $w \in W$, we define $R(w)$  as its bounding rectangle, which marks the space leaders aren't allowed to cross. For a graphical representation see Fig.~\ref*{fig:wbound}.

@@

Since we only allowed the Usage of O- and P-Segments in leaders, the only way for a leader to cross through a line of text is with a P-Segment through whitespace, whereas O-Segments are only usable between lines. Therefore, we can create a routing graph $G=\langle V,E\rangle$ whose Edges $E$ reflect the legal paths a leader can take within $T$.
For each line $l_i \in L$ and each whitespace character $h \in l_i$, one vertex is placed in $s_i$ and $s_{i+1}$ each, located above or below the whitespace character, and with a maximum distance from each of its neighbouring lines. The start and end of each line are also assigned a pair of vertices each. Sites will be represented by the insertion of additional vertices in $S_i$ on a similar height as those next to whitespace of the same line and located directly above the center point of its words $w_j$ bounding rectangle $R(w_j)$. The edges between the vertices are all perfectly horizontal or vertical, and do neither intersect with any bounding rectangle, nor any vertices other than their starting and ending vertex. The resulting graph looks similar to Fig.~\ref*{fig:rgraph}.
In the following, we define vertex $v_a \in s_i$ to be \emph{above} another vertex $v_b \in s_j$, if $i<j$. If $v_a$ and $v_b$ are adjacent to each other, $v_a$ is \emph{directly above} $v_b$. We will also define a vertex $v_b$ to be \emph{between} two vertices $v_a$ and $v_c$, if $v_a, v_b, v_c \in s$ $(s \in S)$ and it is impossible to create a path $P \subseteq s$ using only vertices from $s$ that leads from $v_a$ to $v_c$ without including $v_b$.
Furthermore,we will call the vertices closest to the text area's border to the routing area \emph{Border Vertices} $B \subset V$. They serve as a goal for the depth-first search, and are located at the end of each line in our implementation.

@@

We use a depth-first-search algorithm with the Graph $G$ and the set of sites $V_{ann} \subset V$ as input. The Algorithm prioritizes routing to not yet visited vertices located above the current vertex and terminates either when reaching the right text border or if all possible paths failed to reach the text border, and the resulting SP-path will be monotonically increasing. If the routing for any site $v_{ann} \in V_{ann}$ failed, the algorithm returns $\bot$ for that site. To ensure that routings remain crossing-free, the most recently successfully routed leader $P_{last}$ is used as a ceiling to the available search space. This means that any vertex above the polyline created by the vertices in $P_{last}$ and the horizontal connection of $P_{last}$'s source with the text area's border opposite of the label area is forbidden to use. (See also Fig.~\ref*{fig:ceil}) This path is split into two parts: The path within the graph $(P \subset V)$, leading from the source to the text area's border, and the OPO-Leader that connects the text area's border with the Label's port. The former is implemented as a list of vertices $V_{Path} \subset V$ the leader travels through, whereas the latter is determined by the rightmost vertex of $V_{Path}$, the Port, and the location of the P-Segment. (For an illustration in Pseudocode see Alg.~\ref*{alg:DFS}.)

@@

\begin{lemma}
	The creation of the routing graph is in O(n).
	\label{lem:GraphCreat}
\end{lemma}

\begin{proof}
	As discussed in Section~\ref*{sec:ProbSpec}, each line holds at most $\frac{width}{2}$ words, with an appropriate amount of whitespace in between. Since $width$ is a constant value, the amount of vertices placed is only dependent on the number of lines in the text, therefore $|V|=O(n)$.
	Similarly, each vertex has at most 4 incident edges, representing the two possible placements of P- or O-Segments starting from that vertex, respectively. Therefore, $|E|=O(n)$. As both the creation of new vertices and edges can be done in a constant amount of time, the overall computation time is in $O(n)$.
\end{proof}

For a representation of the graph-generation algorithm in pseudocode, see Alg.~\ref*{alg:GraphCreation}.

@@

\begin{lemma}
	The algorithm only creates crossing-free routings.
	\label{lem:CrosFree}
\end{lemma}
\begin{proof}
	As stated in Section~\ref*{sec:AlgDesc}, we are forbidden to use any vertices above the polyline created by the last successfully routed leader $(P_{last})$ and the direct connection of $P_{last}$'s source and the left text area's border. As this makes it impossible to construct a route that incorporates any vertex above or to the left of $P_{last}$, those two leaders cannot possibly cross each other. This is true for any two consecutive leaders, therefore each leader also cannot cross any of the leaders routed before $P_{last}$, as those are located on the other side of $P_{last}$, which contains only forbidden vertices.
	Therefore, our algorithm produces crossing-free routings.
\end{proof}

For the next part, we first need to define what we consider a \emph{highest legal path}: A \emph{legal path} is one that satisfies both the constraints of monotonicity and being crossing-free, whereas a \emph{highest path} reaches the highest possible border vertex reachable from a given source.

\begin{lemma}
	Our algorithm only returns a highest legal path.
	\label{lem:High}
\end{lemma}
\begin{proof}
	The legality of any generated path is given, due to the algorithm only returning monotonically increasing paths, and Lemma~\ref*{lem:CrosFree}. To show that there are no reachable border vertices above the path's highest reachable border vertex $b_{high}$, we will assume there exists a vertex $b_{top} \in B$ that is located above $b_{high}$ and legally reachable via the path $P_{top}$. As the algorithm prioritizes routing to vertices directly above, the only way for the path to $b_{top}$ to reach higher vertices than $b_{high}$ would be that either $P_{top}$ would generate crossings with previously routed leaders or violate the the monotonicity constraint, which both render $P_{top}$ illegal. Since we also cannot include more than one border vertex in our path, we conclude that such a path cannot exist.
\end{proof}

\begin{theorem}
	The routing problem can be solved with a worst-case time complexity of $O(n^2)$ for all sources.
\end{theorem}
\begin{proof}
	As part of the proof for Lemma~\ref*{lem:GraphCreat}, we have shown that the total number of vertices $|V|$ is in $O(n)$. Since $S \subset V$, $|S|$ must too be in $O(n)$. 
	We have proven our algorithm's solutions to be correct in the Lemmas \ref*{lem:CrosFree} and \ref*{lem:High}.
	The performance of our routing algorithm is easily calculated, as depth-first-search has a worst-case-performance of $|V|+|E|$, which is in $O(n)$ for our problem, as $|E|$ is in $O(n)$ as well. (See Lemma~\ref*{lem:GraphCreat})
	Therefore, our overall computation time comes out to $|S| \times O(n)$, or $O(n^2)$.
\end{proof}

\chapter{Implementation}
The program was written in Java 1.8.0u40, using JGraphT 1.0.1\footnote{\url{http://jgrapht.org/}} as graph library. No other external libraries were used.
The program itself can be found at GitHub \footnote{\url{https://github.com/e1125755/AnnotationRouting}}.

\section{Architecture}
The program is divided into three parts. The GUI and program window only propagate user interaction to other parts of the program. It will be discussed further in the next section.
The main drawing area handles the visualization of the results, calculates where to place the text on-screen, and also generates the routing graph, as this can be done alongside the text placement.
\change{The routing is done in a separate class, and can easily be exchanged with another algorithm because of this.}
\section{GUI}
\label{sec:GUI}

The GUI of the program is kept simple, as the main goal was to demonstrate the routing algorithm. All customization options are located on the control panel above the drawing area, which contains the text, leaders and annotations. (See also Fig.~\ref*{fig:ProgWindow})

The control panel's first three options determine the font the program uses for the main text, whether it is bold or italicized and its size. \change{The annotations' text currently uses a smaller version of this font. The fourth and last option allows the selection of different routing algorithms, although the current options are mostly variations on the routing algorithm described in Section~\ref*{sec:AlgDesc}.} After changing any value, the program will automatically be notified of the changes and react accordingly.

\begin{figure}%TODO: Replace with better screenshot!
	\captionsetup{justification=centering, margin=0.75cm}
	\centering
	\includegraphics[scale=0.75]{ProgramWindowAnnotated.png}
	\caption{Picture of the routing program}
	\label{fig:ProgWindow}
\end{figure}

\section{Drawing Area}
\change{As this program is a proof of concept, the drawing area itself can't be interacted with, it only serves to display the algorithm's results. If we were to adapt this program for regular use, the user should be able to manually change routings or move labels around here.
Due to reasons described below in Section~\ref*{sec:Challenge}, each word has to be placed individually, which allows us to create the routing graph alongside the darwing process. 
After the text has been drawn completely, the routing algorithm is invoked from here, and its results are visualized.}

\section{Routing Algorithm}

\change{The Routing algorithm has already been described thoroughly in Section~\ref{sec:AlgDesc}, that process is implemented by creating ordered lists of the vertices traversed by the SP-Path, along with some additional information on where to draw the P-segment located in the Routing Area. While the algorithm offers methods to route each annotation individually, we currently route all annotations at the same time, which vastly decreases the required communication between the algorithm and the rest of the program.}

\section{Challenges}
\label{sec:Challenge}
Displaying the text on-screen without losing the position of the whitespace characters turned out to be harder than expected. While Java offers a class that divides any string of characters into separate lines, the resulting line object does not allow for it's text to be extracted, which makes it unusable for our purposes. This problem was solved by placing each word on the screen individually. For simplicity's sake, we only allowed line breaks between words, whenever a whitespace character was encountered.

Another issue was the visualization of the routing graph and the leaders. While JGraphT offers ways to display graphs, it wasn't possible to influence the placement of individual vertices. Our solution was forgoing these options and drawing the graph \change{unassisted}, assigning coordinates to each vertex on creation. Vertices are represented by small rectangles, whereas edges are represented with straight lines. To avoid confusion between crossing leaders and leaders that incorporate the same vertex in their path, but curve away from each other, all leaders are drawn with rounded corners, which can be seen in Fig.~\ref{fig:ProgWindow}.

@@

When the program starts, it first scans the text word by word in reading order, placing each word $w \in W$  on-screen, and saving the content of each encountered annotation for later. As the words are placed, vertices are added to the graph after each word, and at the start and end of each line. Source vertices are inserted whenever an annotation is encountered, and contain a reference to their annotation in addition to the information contained in regular vertices.

@@

Once the graph has been generated, a list of all source vertices in reading order is passed to the selected routing algorithm, which then determines leader routing and annotation placement for each of them and returns the results to the main program, which then draws the results. (For details on the routing algorithm, see section~\ref*{sec:AlgDesc}.)

If any of the parameters on the control panel (see section~\ref*{sec:GUI}) are changed, the process re-starts from the beginning.

%TODO: Schematischen Graph (Flowchart?) von oben beschriebenem Ablauf erstellen!
%As the Graph's vertices represent fixed locations on the canvas, they each have coordinates associated with them, with the vertices representing the sites containing extra information, such as references to the corresponding annotation and to the leader connecting the two (if existent).
%The additional information stored in the vertices is used for both routing and drawing the leaders: The routing was done via a depth-first search algorithm that prioritized vertices located further up if possible, and vertices to the right otherwise. After the algorithm terminates, it returns  a vertex-based graph walk describing the route from the site to the border of the text area alongside information about where to draw the OPO-segment in the buffer zone if successful, or a walk containing only the starting vertex, if no valid path was found. Using this information alongside the positional data stored in the vertices, the polyline representing the leader can be drawn.

%TODO: Schönheitsfeatures, etc.
